---
title: "Project_1_wordcloud"
author: "WendyD"
date: "9/21/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, results="hide", message=FALSE}
library(tibble)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(patchwork)
library(RColorBrewer)
library(gridExtra)
library(data.table)
library(tm)
library(SnowballC)
library(wordcloud)
```

```{r setup, echo=FALSE}
library(knitr)
opts_knit$set(root.dir=normalizePath("../"))
```

# Introduction

This report, hopefully, will help you understand what make up the majority of Democratic and Republican voting base, their preference, what they like about their parties of choice. Some other criteria that this report covers including: education background, .....(ADD ON LATER)

The data I used for this report comes from American National Election Studies(ANES). The original data has 1029 variables. To make thing easier, I subseted the data by variables of interest, and saved it as data1

# What make a person choose one party over the other?

Since the day I came to the US, I have always been intrigued about the US democracy system. Coming from a Communist country, where the Vietnam Communist Party is the sole governing political party, I found the rivalry between the Democratic and Republican parties very interesting. With so many options, I have always wondering: What make a person choose one party over another? 

I looked into their answers to the questions: what do voters like/don't like about the party? I analyzed their answers, and made a word cloud plots based on the frequency that key words popped up. The plots gave some interesting insights on the each party's characteristics. 


From what the word clouds show, it appears the top common reason that make voters hate one party is Domestic policies. The words "domestic policies", "philosophy" , "party character" showed up on both democratic and republican graph. This is easy to understand, as people have different opinions on how their countries should be run, as well as their philosophies. One interesting thing that we see is  people who disagree with the democratic party' management skill. This does not happen with  the republican party. Apparently, voters across the board think that the democratic party lack credential in their management skills. On the other hand, Group connection quality is among the top reasons why voters don't like republican party.

# Republican vs Democratics: which party has a "smarter" voting base?

Graduated from a liberal art college in dominantly blue state, Minnesota, I have always believe that people who support Democratics are smarter, more educated, hence the liberal points of views, in comparison to Republican supporters. At least, that is what my American classmates told me. I wonder if that is true. 

To answer this question, I take a look at the education distribution of the voters versus political identification. I started with 2016, which to many people, had the most unexpected election outcomes. 

```{r}
data1 <- read.csv("output/anes_edu.csv", header=TRUE)
```


```{r, echo=FALSE}
data2 <- data1[ data1$year == 2016,  ]
a<- t((data2[data2$partyID == 1,] %>% count(eduID)*100/sum(data2$partyID == 1))[2])
b<- t((data2[data2$partyID == 3,] %>% count(eduID)*100/sum(data2$partyID == 3))[2])
edubyparty<-rbind(a,b)
edubyparty<-as.data.frame(edubyparty)
names(edubyparty)<- c("Grade School", "High School", "Some Colleges", "College or higher")
rownames(edubyparty)<-c("Democratics", "Republican")
print(edubyparty)
```

It seems like so far in 2016, people who have college degrees and higher tend to identify themselves as democrat, though the ratio is not much higher.

Let us look at the data across the year, to see if this phenomenon still apply over time

```{r, echo=FALSE}
# histogram time series to plot
frequency <- data1 %>% count(year, eduID, partyID)

# add new col number of people per party, rescale to fit frequency
partySum <- (data1 %>% count(year, partyID))
partySum.expanded <- partySum[ rep(row.names(partySum), each=4) , 2:3 ]

# sort by frequency file by partyID
frequency <- frequency[ order(frequency$year, frequency$partyID) , ]

# then add new column of partySum
frequency[, 5] <- partySum.expanded[, 2]                        

# add new col of percentage
frequency[, 6] <- frequency[, 4] / frequency[, 5] * 100

# change labels
names(frequency) <- c("year", "eduID", "partyID", 
                      "count", "partySum", "percentage" )

# change eduID labels
eduLabel <- data.frame( 
    ID = c(1,2,3,4), 
    level = c("1 Grade School", "2 Highschool", "3 Some College", "4 College")
)

frequency$eduID <- eduLabel$level[ match(frequency$eduID, eduLabel$ID) ]

# graph
ggplot( 
    data=frequency, 
    aes(
        x=year, 
        y=percentage, 
        colour=as.factor(partyID)
        )
) + 
geom_point() + 
geom_line() +
facet_wrap(frequency$eduID ~ .) + 
scale_color_manual(
    name="Party Identification",
    labels=c("Dem", "Ind", "Rep"),
    values=c("blue","brown","red")
) +
scale_x_continuous( limits=c(1948,2020) )
```

Contrary to my beliefs, 2016 is an anomaly.It seems that across the years, Democrats voter base fall mostly into the two categories: (`Grade school` and `High School`), while a Republican will more likely in the other two (`Some colleges` and `College or higher`).

Across the years, there has been a decline in voters without a college degree. College-educated voters, on the other hand, have increased, according to the analysis. There have been more and more people went to get higher degrees, thus introducing an upward trends in some college and colleges education groups on the graph. 

This realization came to me as a total surprise.  To render my analysis more quantitative, I look at percentage difference between the Democrats and Republicans for each of the educational groups

```{r, echo=FALSE}

# count by eduID only
dataBox <- (data1 %>% count(year, eduID))
# Make two new columns of total Dem and Rep
dataBox[, 4] <- (data1[ data1$partyID == 1, ] %>% count(year, eduID))$n
dataBox[, 5] <- (data1[ data1$partyID == 3, ] %>% count(year, eduID))$n
# input total number of dem and rep
foo5 <- (data1 %>% count(year, partyID))$n
foo6 <- as.data.frame(t(matrix(foo5, nrow=3, ncol=length(foo5)/3)))
foo6.expanded <- foo6[ rep(row.names(foo6), each=4) , c(1,3) ]
dataBox[, 6:7] <- foo6.expanded
# probability
dataBox[, 8] <- dataBox[, 4] / dataBox[, 6] * 100
dataBox[, 9] <- dataBox[, 5] / dataBox[, 7] * 100
# probability difference
dataBox[, 10] <- dataBox[, 8] - dataBox[, 9]


names(dataBox) <- c("year", "eduID", "tot", "dem", "rep",
                 "demTot", "repTot", "P(dem)", "P(rep)", "spread")

# plot
dataBox$eduID <- as.factor(dataBox$eduID)
ggplot(dataBox, aes(x=eduID, y=spread))+
        xlab("Education groups")+
        ylab("Percentage")+
        geom_boxplot(aes(factor(eduID), spread))+
        scale_x_discrete(labels = c("Grade school", "High school", "Some College", "College degree"))

```

This boxplot indicates two things:

- People with high school degree or lower tend to vote for democrats more. On the other hands, Republicans have a stronger base in higher educational groups (college degree or higher). 
-  2016 is an outlier for the `college and higher` category, while 2012 is an outlier for the `some colleges` category. This means that there was a shift in political preference among the higher educational groups in 2016, as more people with college degree vote for democrats more than for republicans.  

# Keywords about each party

I tried to reason the political preference for each different educational background by looking at voter answers to the questions: what do voters like/don't like about the party? I analyzed their answers, and made a word cloud plots based on the frequency that key words popped up. The plots gave some interesting insights on the each party's characteristics. 

```{r, echo=FALSE}
data1<-read.table("output/anes_trimmed.csv", header = TRUE, sep=",")
```


```{r, echo=FALSE}
likedem<-data1$VCF0375b
likedem2<-data1$VCF0376b
likedem3<-data1$VCF0377b
likedem4<- data1$VCF0378b
likedem5<- data1$VCF0379b

likedem<-as.data.frame(likedem)
colnames(likedem)<-"LikeID"

likedem2<-as.data.frame(likedem2)
names(likedem2)<-"LikeID"

likedem3<-as.data.frame(likedem3)
names(likedem3)<-"LikeID"

likedem4<-as.data.frame(likedem4)
names(likedem4)<-"LikeID"

likedem5<-as.data.frame(likedem5)
names(likedem5)<-"LikeID"

likedem_total<-rbind(likedem,likedem2,likedem3, likedem4, likedem5)
likedem_total<-likedem_total[!is.na(likedem_total)]
likedem_total<-as.data.frame(likedem_total)
names(likedem_total)<- "LikeID"

#create data frame used for label
like_label <- data.frame(
        ID = c(11, 12,21,22,23,24,31,32,33,34,35,40,50), 
        Reason = c("PeopleinParty", 
                   "PartyCharacter", 
                   "Candidate Ability", 
                   "Leadership Qualities",
                   "Candidate Personal Qualities", 
                   "Party Connection", 
                   "Management", 
                   "Philosophy", 
                   "DomesticPolicy", 
                   "F-Policies",
                   "GroupConnect", 
                   "Etc",
                   "Events Unique to One campaign"
        )
)

### Match label with the LikeID
likedem_total$LikeID <- like_label$Reason[match(likedem_total$LikeID, like_label$ID)]

### CREATE WORD CLOUD PLOT

text3<-likedem_total$LikeID
wc <- count(likedem_total, LikeID)
wordcloud(words=wc$LikeID, freq = wc$n, scale=c(2,.5), random.order = FALSE,rot.per=.5)

```

```{r,echo=FALSE}
likerep<-data1$VCF0387b
likerep2<-data1$VCF0388b
likerep3<-data1$VCF0389b
likerep4<- data1$VCF0390b
likerep5<- data1$VCF0391b

likerep<-as.data.frame(likerep)
names(likerep)<-"LikeID"

likerep2<-as.data.frame(likerep2)
names(likerep2)<-"LikeID"

likerep3<-as.data.frame(likerep3)
names(likerep3)<-"LikeID"

likerep4<-as.data.frame(likerep4)
names(likerep4)<-"LikeID"

likerep5<-as.data.frame(likerep5)
names(likerep5)<-"LikeID"

likerep_total<-rbind(likerep,likerep2, likerep3, likerep4, likerep5)   
likerep_total<-likerep_total[!is.na(likerep_total)]
likerep_total<-as.data.frame(likerep_total)
names(likerep_total)<- "LikeID"

likerep_total$LikeID <- like_label$Reason[match(likerep_total$LikeID, like_label$ID)]

## Plotting
text4<-likerep_total$LikeID
wc2 <- count(likerep_total, LikeID)
wordcloud(words=wc2$LikeID, freq = wc2$n, scale=c(2,.5), random.order = FALSE,rot.per=.5)


```

Democratic party win voters favor with their Group Connection quality, and their Domestic Policies. Voters, on the other hand, resonance with Republican Party on their Domestic Policy, Philosophy, Management skills. 

We repeat the same process to answer the question: what characteristics in each parties that make voters hate: 


```{r, echo=FALSE}
# ATTEMPT DISLIKE about DEMOCRATIC PARTY
dislikedem<-data1$VCF0381b
dislikedem<-dislikedem[!is.na(dislikedem)]
dislikedem<-as.data.frame(dislikedem)
names(dislikedem)<-"DislikeID"

#VCF0382
dislikedem3<-data1$VCF0382b
dislikedem3<-dislikedem3[!is.na(dislikedem3)]
dislikedem3<-as.data.frame(dislikedem3)
names(dislikedem3)<-"DislikeID"

## VCF0383
dislikedem6<-data1$VCF0383b
dislikedem6<-dislikedem6[!is.na(dislikedem6)]
dislikedem6<-as.data.frame(dislikedem6)
names(dislikedem6)<-"DislikeID"

## VCF0384
dislikedem5<-data1$VCF0384b
dislikedem5<-dislikedem5[!is.na(dislikedem5)]
dislikedem5<-as.data.frame(dislikedem5)
names(dislikedem5)<-"DislikeID"

# rbind and make new data frame
dislikedem_total<-rbind(dislikedem, dislikedem3, dislikedem6, dislikedem5)
dim(dislikedem_total)

# match the dislikeID with answer key
dislike_label1 <- data.frame(
        ID = c(11, 12,21,22,23,24,31,32,33,34,35,40,50), 
        Reason = c("PeopleinParty", 
                 "PartyCharacter", 
                 "Candidate Ability", 
                 "Leadership Qualities",
                 "Candidate Personal Qualities", 
                 "Party Connection", 
                 "Management", 
                 "Philosophy", 
                 "DomesticPolicy", 
                 "F-Policies",
                 "GroupConnect", 
                 "Etc",
                 "Events Unique to One campaign"
                 )
        )

dislike_label2 <- data.frame(
        ID= c(1,2,3,4,5,6,7,8,9,10), 
        Reason= c("PeopleinParty",
                  "Management", 
                  "Philosophy", 
                  "DomesticPolicy", 
                  "DomesticPolicy", 
                  "ForeignPolicy", 
                  "Bad for groups of interest", 
                  "Good for group of interest", 
                  "Party Attitude",
                  "Others"
                  )
        )
dislike_label<-rbind(dislike_label1, dislike_label2)

#match label with factor in main data frame
dislikedem_total$DislikeID <- dislike_label$Reason[
        match( dislikedem_total$DislikeID, dislike_label$ID )
        ]

####### CREATE WORD CLOUD PLOT

text<-dislikedem_total$DislikeID
cd2 <- count(dislikedem_total, DislikeID)
wordcloud(words=cd2$DislikeID, freq = cd2$n,scale=c(2,.5), random.order = FALSE,rot.per=.5)


```

```{r, echo=FALSE}
dislikerep<-data1$VCF0393b
dislikerep2<-data1$VCF0394b
dislikerep4<- data1$VCF0395b
dislikerep5<- data1$VCF0396b
dislikerep6<- data1$VCF0397b

# create dataframe and rename column
dislikerep<-as.data.frame(dislikerep)
names(dislikerep)<-"DislikeID"

dislikerep2<-as.data.frame(dislikerep2)
names(dislikerep2)<-"DislikeID"

dislikerep4<-as.data.frame(dislikerep4)
names(dislikerep4)<-"DislikeID"

dislikerep5<-as.data.frame(dislikerep5)
names(dislikerep5)<-"DislikeID"

dislikerep6<-as.data.frame(dislikerep6)
names(dislikerep6)<-"DislikeID"

dislikerep_total<-rbind(dislikerep, dislikerep2,dislikerep4,dislikerep5,dislikerep6)
dislikerep_total<-dislikerep_total[!is.na(dislikerep_total)]
dislikerep_total<-as.data.frame(dislikerep_total)
names(dislikerep_total)<- "DislikeID"


### Match label with the dislikeID
dislikerep_total$DislikeID <- dislike_label$Reason[match(dislikerep_total$DislikeID, dislike_label$ID)]

### CREATE WORD CLOUD PLOT

text2<-dislikerep_total$DislikeID
cd <- count(dislikerep_total, DislikeID)
wordcloud(words=cd$DislikeID, freq = cd$n, scale=c(2,.5), random.order = FALSE,rot.per=.5)


```

From what the word clouds show, it appears the top common reason that make voters hate one party is Domestic policies. The words "domestic policies", "philosophy" , "party character" showed up on both democratic and republican graph. This is easy to understand, as people have different opinions on how their countries should be run, as well as their philosophies. One interesting thing that we see is  people who disagree with the democratic party' management skill. This does not happen with  the republican party. Apparently, voters across the board think that the democratic party lack credential in their management skills. On the other hand, Group connection quality is among the top reasons why voters don't like republican party.

Connecting this to our previous finding, that higher educated voters prefer republicans, the word cloud plots make total sense. People who have lower income tend to come from a less educated background. These people choose Democrats because the party offer a superior group connection quality, with a promise of equal pay, higher tax for the reach, and most importantly, inclusivity for the poor. The voters who have higher education, enjoyed a higher pay positions, want to maintain their social status, so they resonance more with the Republican's conservative philosophy. These voters place an emphasis on the party's management skill in preserving the rich's benefits. 

# Conclusion


